---
title: "Getting Started"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

The goal of champsmortality is to provide functions for calculating factor-adjusted mortality fractions and rates at CHAMPS sites and catchments for various causes.

## Installation

You can install the development version of champsmortality with the following:

``` r
install.packages("remotes") # one time only
remotes::install_github("ki-tools/champs-mortality")
```

Once installed, load the library:

```{r}
library(champsmortality)
```

### Data setup

The first time you use this package, you need to place the appropriate data files in a data directory that the package will pull from to perform the calculations. A function `create_dataset_directory()` is provided to help get this set up.

Here we will create the dataset directory in a temporary location. 

```{r strip.white=FALSE}
data_dir <- tempfile()
create_dataset_directory(data_dir)
```

As long as you maintain the specified directory on your computer, you do not need to run this command again. We recommend choosing a persistent directory.

This creates the directory and copies some data files over that come with the package:

```{r}
list.files(data_dir)
```

These files come with the package. They will be described in greater detail in the [next article](datasets.html).

The `config.yaml` file contains pointers to each dataset that is required for analysis:

```{r}
cat(paste(readLines(file.path(data_dir, "config.yaml")), collapse = "\n"))
```

Each entry specifies the file name of each required data file. Note the three datasets that already exist and an additional 5 that don't come with the package due to their sensitive nature. These datasets can be obtained from [CHAMPS](https://champshealth.org/data/) and more information about their contents and structure will be provided in the [next section](datasets.html).
